---
title: "Decision Tree Herpes Virus Encephalopathy"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Decision Tree Herpes Virus Encephalopathy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This is an example that we use to illustrate decision trees in our DARTH teaching materials.  The example involves the treatment of herpes simplex ecncephalopathy (HVE) by either treating everyone without biopsy (Treat), not treating or taking biopsy (DoNotTreat), and Taking biopsy and treat if biopsy posivite (Biopsy). 

We will first define the gmod model and the parameters and then we will show how one would code the model direclty in R and compare the outcomes. 

# GMOD
```{r setup}
rm(list = ls())
library(gmod)
```

gmod model definition
p_comp() takes in two arguments, one is the special keyword decision, and the other is a custom one.  This is a good example that custom arguments and values can be passed to define functions for the event_mapping and the payoffs. Thus, the return values can be a function of these arguments and it can reduce the amount of typing required if one choses to define multiple functions. 

```{r}
mygmod <- gmod() + 
  decisions("DoNotTreat", "Treat", "Biopsy") + 
  #final_outcomes("Death", "HVE_comp", "no_HVE_comp", "OVE_comp", "no_OVE_comp") +
  #events("DIE", "HVE","get_comp") + 
  event(name = "DIE",  
            scenarios = c(T, F), 
            probs = c(pDie(decision), Inf), 
            outcomes = c("Death", "HVE_event")) + 
  event(name = "HVE_event",  
            scenarios = c(T, F), 
            probs = c(p_HVE, Inf), 
            outcomes = c("get_HVE_comp", "get_OVE_comp")) +
  event(name = "get_HVE_comp", 
            scenarios = c(T, F),
            probs = c(p_comp(decision, HVE = TRUE), Inf),
            outcomes = c("HVE_comp", "no_HVE_comp"))  +
  event(name = "get_OVE_comp", 
            scenarios = c(T, F),
            probs = c(p_comp(decision, HVE = FALSE), Inf),
            outcomes = c("OVE_comp", "no_OVE_comp")) + 
  #payoffs(cost = cost(decision, final_outcome, prop_with_event("HVE"=TRUE, decision)), 
  payoffs(cost = cost(decision, final_outcome),  
          effectiveness = effectiveness(decision, final_outcome))

```


Define the model parameters. 

```{r}
v_names_str    <- c("No Tx", "Tx All", "Biopsy")  # names of strategies
n_str          <- length(v_names_str)             # number of strategies
wtp            <- 100000                          # willingness to pay threshold

params <- list(
        wtp            = 100000 ,                         # willingness to pay threshold

        # Probabilities,
        p_HVE          = 0.52   ,# prevalence of HVE
        p_HVE_comp     = 0.71   ,# complications with untreated HVE
        p_OVE_comp     = 0.01   ,# complications with untreated OVE
        p_HVE_comp_tx  = 0.36   ,# complications with treated HVE
        p_OVE_comp_tx  = 0.20   ,# complications with treated OVE
        p_biopsy_death = 0.005  ,# probability of death due to biopsy

        # Costs,
        c_VE           = 1200   ,# cost of viral encephalitis care without complications
        c_VE_comp      = 9000   ,# cost of viral encephalitis care with complications
        c_tx           = 9500   ,# cost of treatment
        c_biopsy       = 25000  ,# cost of brain biopsy

        # QALYs,
        q_VE           = 20     ,# remaining QALYs for those without VE-related complications
        q_VE_comp      = 19     ,# remaining QALYs for those with VE-related complications
        q_loss_biopsy  = 0.01   ,# one-time QALY loss due to brain biopsy
        q_death_biopsy = 0      # remaining QALYs for those who died during biopsy
)

```

define custom user functions
```{r}
pDie <- function(decision){
  if (decision == "Biopsy") p_biopsy_death else 0
}

p_comp <- function(decision, HVE){
  if (decision == "DoNotTreat" & HVE ) return(p_HVE_comp)
  if (decision == "DoNotTreat" & !HVE )return(p_OVE_comp)
  if (decision == "Treat" & HVE ) return(p_HVE_comp_tx)
  if (decision == "Treat" & !HVE ) return(p_OVE_comp_tx)
  if (decision == "Biopsy" & HVE) return(p_HVE_comp_tx)
  if (decision == "Biopsy" & !HVE) return(p_OVE_comp)
}

c_HVE <- function(decision){
  if (decision == "biopsy") c_tx else 0
}

cost <- function(decision, final_outcome){ 
  c_biopsy*(decision=="Biopsy") + 
    c_tx*(decision=="Treat" | (decision=="Biopsy" & final_outcome %in% c("HVE_comp", "no_HVE_comp"))) + 
    c_VE_comp*(final_outcome %in% c("HVE_comp", "OVE_comp")) + 
    c_VE*(final_outcome %in% c("no_HVE_comp", "no_OVE_comp")) 
}

effectiveness <- function(decision, final_outcome){
  -q_loss_biopsy*(decision=="Biopsy") + 
    q_VE_comp*(final_outcome %in% c("HVE_comp", "OVE_comp")) + 
    q_VE*(final_outcome %in% c("no_HVE_comp", "no_OVE_comp"))
}
```

testing some of the functions outside of the model 
```{r}
#p_comp(decision = "Biopsy", HVE = FALSE)
#
#cost("Biopsy", "HVE_comp")
#effectiveness("Biopsy", "Death")
```

```{r}
model_struc <- gmod_build(mygmod)
model_struc
```

```{r}
model_res <- gmod_gen_model_function(model_struc)
model_res
```

```{r}
my_decision_model(params)
```

# Direct coding in R
And here is what the code looks like if one wants to code the transitions all in R directly without using gmod
```{r}
# Create vector of weights for each strategy 
  
v_w_no_tx  <- c(    p_HVE  *      p_HVE_comp     ,  # HVE, complications
                    p_HVE  * (1 - p_HVE_comp)    ,  # HVE, no complications
               (1 - p_HVE) *      p_OVE_comp     ,  # OVE, complications
               (1 - p_HVE) * (1 - p_OVE_comp))      # OVE, no complications
  
v_w_tx     <- c(    p_HVE  *      p_HVE_comp_tx  ,  # HVE w/tx, complications
                    p_HVE  * (1 - p_HVE_comp_tx) ,  # HVE w/tx, no complications
               (1 - p_HVE) *      p_OVE_comp_tx  ,  # OVE w/tx, complications
               (1 - p_HVE) * (1 - p_OVE_comp_tx))   # OVE w/tx, no complications
  
v_w_biopsy <- c(p_biopsy_death                   ,  # biopsy death
               # no biopsy death.,   HVE w/tx,        complications
               (1-p_biopsy_death)   *    p_HVE  *    p_HVE_comp_tx  ,  
               # no biopsy death.,   HVE w/tx,     no complications
               (1-p_biopsy_death)   *    p_HVE  * (1-p_HVE_comp_tx) ,  
               # no biopsy death.,        OVE,        complications
               (1-p_biopsy_death)   * (1-p_HVE) *      p_OVE_comp   ,  
               # no biopsy death.,        OVE,     no complications
               (1-p_biopsy_death)   * (1-p_HVE) * (1 - p_OVE_comp))      
  
# Create vector of final_outcomes (QALYs) for each strategy 
  
v_qaly_no_tx  <- c(q_VE_comp ,          # HVE, complications
                   q_VE      ,          # HVE, no complications
                   q_VE_comp ,          # OVE, complications
                   q_VE)                # OVE, no complications
  
v_qaly_tx     <- c(q_VE_comp ,          # HVE, complications
                   q_VE      ,          # HVE, no complications
                   q_VE_comp ,          # OVE, complications
                   q_VE)                # OVE, no complications
  
  
v_qaly_biopsy <- -q_loss_biopsy     +   # loss due to biopsy
                  c(q_death_biopsy  ,   # biopsy complications
                    q_VE_comp       ,   # no biopsy comp., HVE w/tx, complications 
                    q_VE            ,   # no biopsy comp., HVE w/tx, no complications
                    q_VE_comp       ,   # no biopsy comp., OVE, complications
                    q_VE)               # no biopsy comp., OVE, no complications
  
# Create vector of costs for each strategy 
  
v_cost_no_tx  <- c(c_VE_comp ,          # HVE, complications
                   c_VE      ,          # HVE, no complications
                   c_VE_comp ,          # OVE, complications
                   c_VE)                # OVE, no complications
  
v_cost_tx     <- c_tx +                 # cost of treatment
                 c(c_VE_comp ,          # HVE, complications
                   c_VE      ,          # HVE, no complications
                   c_VE_comp ,          # OVE, complications
                   c_VE)                # OVE, no complications
  
v_cost_biopsy <- c_biopsy           +   # cost of biopsy procedure
                 c(0                ,   # cost of death (zero)
                   c_VE_comp + c_tx ,   # no biopsy comp., HVE w/tx, complications 
                   c_VE + c_tx      ,   # no biopsy comp., HVE w/tx, no complications
                   c_VE_comp        ,   # no biopsy comp., OVE, complications
                   c_VE)                # no biopsy comp., OVE, no complications
 
# Calculate total utilities for each strategy 
total_qaly_no_tx  <- v_w_no_tx  %*%  v_qaly_no_tx      
total_qaly_tx     <- v_w_tx     %*%  v_qaly_tx
total_qaly_biopsy <- v_w_biopsy %*%  v_qaly_biopsy
  
# Calculate total costs for each strategy 
total_cost_no_tx  <- v_w_no_tx  %*%  v_cost_no_tx    
total_cost_tx     <- v_w_tx     %*%  v_cost_tx
total_cost_biopsy <- v_w_biopsy %*%  v_cost_biopsy
  
# vector of total QALYs
v_total_qaly <- c(total_qaly_no_tx, total_qaly_tx, total_qaly_biopsy) 
# vector of total costs
v_total_cost <- c(total_cost_no_tx, total_cost_tx, total_cost_biopsy) 
# calculate vector of nmb
v_nmb        <- v_total_qaly * wtp - v_total_cost                      
  
# Name final_outcomes
names(v_total_qaly) <- v_names_str  # names for the elements of the total QALYs vector
names(v_total_cost) <- v_names_str  # names for the elements of the total cost vector
names(v_nmb)        <- v_names_str  # names for the elements of the nmb vector
  
df_output <- data.frame(Strategy =  v_names_str,
                        Cost     =  v_total_cost,
                        Effect   =  v_total_qaly,
                        NMB      =  v_nmb)

# model output
df_output

```

