---
title: "Introduction to Decision Trees"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to Decision Trees}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

gmod is build on the Grammar of Modeling, and meant to allow for building decision trees and Markov models using a concise and flexible syntax.  This introductory example illustrates a simple decision tree that we will gradually build on in other vignettes.

If you haven't already, install gmod using devtools.  Also, make sure to enable use_vignettes so that vignettes render correctly.
```{r}
#library(devtools)
#install_github("hjalal/gmod", use_vignettes = TRUE, force = TRUE)
```

```{r setup}
rm(list = ls())
library(gmod)
```

First, we define the gmod object by defining its layers.  
```{r}
mygmod <- gmod() + 
  decisions("A", "B") + 
  event(name = "get_infection",  
        results = c(T, F), 
        probs = c(pSick(decision), Inf), 
        outcomes = c("Sick", "Healthy")) + 
  payoffs(life_expectancy = compute_le(final_outcome))

```
The `gmod()` function creates a gmod object of type "Decision" standing for a decision tree.  
For a Markov model, please see the Markov model Vignette.  
The `decision()` function describes the decision options.  These can later be referred to with 
the special keyword (decision) below.  There should be only one decisions layer per gmod object.
The `event_mapping()` describes a single event "get_infection". This event takes two values TRUE or FALSE. [Below we expand this to use more than two values.]  If TRUE the person ends up with the "Sick" final_outcome, otherwise they end up with the "Healthy" final_outcome. The probs argument defines the probability of getting sick.  Notice that we are using an external function `pSick` (that we will define next). This function returns a different value depending on the decision. The `Inf` is a place holder computes the complementary probbility (i.e., `1-pSick()`) and reduces the amount of typing required. 
The payoffs layer defines the payoffs.  Here we use a single payoff life_expectancy the value of which is computed using the `compute_LE()` function by taking the special keyword final_outcome defined below. final_outcomes are the outcomes of the final events in the gmod object.  

```{r}
pSick <- function(decision){
  if (decision == "A") 0.5 else 0.7
}
```
pSick is a simple function here that will return 0.5 if the decision="A", otherwise 0.7.  Later we will show how to create more complex probabilities that can take into account the value of prior events which will help us keep the code concise.

```{r}
compute_le <- function(final_outcome){
  if (final_outcome == "Sick") 20 else 40
}
```
Similar to `pSick` above, we define the `compute_le()` as a function of the final_outcome.  If Sick, it returns 20 years, if Healthy, it returns 40 years. 

NOTE: R is case sensitivie, so you have to make sure that all the oucomes, payoffs, values, decisions follow the same case.  For example, if we use `final_outcome=="sick"` that is not the same as `final_outcome=="Sick"`. 

Then, we can build the gmod object by calling `gmod_build()` function which returns the formulae of our decision tree as a list 
```{r}
model_struc <- gmod_build(mygmod)
model_struc
```

```{r}
gmod_gen_model_function(model_struc)
```

```{r}
my_decision_model(model_struc)
```


This list returns information about our tree.  For example, the final_outcomes and the payoffs. It also returns two tables the first one lists the expanded form of the tree and the formulae by decision and final_outcome and each path that leads to the final_outcome.  The second is a summary table for the formulae involved in the payoffs.  Let's take a look at the payoff summary table.
```{r}
model_struc$summary_formulae
```
This table will return two formuale for life expectancy, one per decision. For example life expectancy of A equals `(1-(pSick('A')))*compute_le('Healthy')+(pSick('A'))*compute_le('Sick')`.  This is a meaningful formula as it describes the life expectancy of decision A as a weighted sum of probability of getting sick given A and life expectancy given each final_outcome Sick and Healthy.  We can also evaluate this formula directly of gmod

```{r}
(1-(pSick('A')))*compute_le('Healthy')+(pSick('A'))*compute_le('Sick')
```
This returns 30 years on average for treatment A.

We can formally evaluate our gmod object by calling the function `gmod_evaluate()`:
```{r}
model_res <- gmod_evaluate(model_struc)
model_res
```

The main output is a table containing the values of the payoffs.  We can see the 30 years for A, and also we can see that the life expectancy for B is 26 years.  Which we can double check by examing the formula for B:
```{r}
(1-(pSick('B')))*compute_le('Healthy')+(pSick('B'))*compute_le('Sick')
```

In other vignette, we show other features, including:

1. Event final_outcomes other than TRUE and FALSE
2. Multiple sequential events
3. Multiple payoffs (e.g., costs and effectiveness) for cost-effectiveness analyses
4. The probability of an event depending on the final_outcome of upstream events
5. The payoffs depending on the intermediate final_outcomes from upstream events
