---
title: "M1 Introduction to Markov modeling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{M1 Introduction to Markov modeling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette replicates our introductory tutorial for Markov models in *Medical Decision Making*:  

Alarid-Escudero, F., Krijkamp, E., Enns, E. A., Yang, A., Hunink, M. M., Pechlivanoglou, P., & Jalal, H. (2023). An introductory tutorial on cohort state-transition models in r using a cost-effectiveness analysis example. Medical Decision Making, 43(1), 3-20.

If you haven't already, install gmod using devtools.  Also, make sure to enable use_vignettes so that vignettes render correctly.
```{r}
#library(devtools)
#install_github("hjalal/gmod", use_vignettes = TRUE, force = TRUE)
```

```{r setup}
rm(list = ls())
library(gmod)
```

The Markov model involves a cost-effectiveness example with 4 decisions and 4 states. For more details on the model, please refer to the reference above. First we define the gmod object and the layers. 
```{r}
gmod(name = "mygmod") + 
  decisions("StandardOfCare", "StrategyA", "StrategyB", "StrategyAB") + 
  states(names = c("H", "S1", "S2", "D"), init_probs = c(1,0,0,0)) +  
  event(name = "get_event",  
                scenarios = c("recover", "getsick", "progress", "die", "stay"), 
                probs = c(pRecover(state), pGetSick(state), pProgress(state, decision), pDie(state), Inf), 
                outcomes = c("H", "S1", "S2", "D", "curr_state")) +  
  payoffs(cost = cost(state, decision), 
          utility = utility(state, decision),
          discount_rates = c(0.03, 0.03))
```

The `gmod()` function defines the model type as a Markov model and number of cycles as 75. The `decision()` function defines the treatment strategies.  The `initial_probs()` defines the initial probabilities. Here we only define the "Moderate" states as the starting state with probability equals 1. The other states will be set to 0. 

Since the tutorial assumes that all the events happen simultaneously (i.e., no sequence), we add a single `event_mapping()` layer we model the individual final_outcomes separately.  We can interpret the event_mapping as the following: The event `get_event` can have 5 values: recover, getsick, progress, die, or stay in the same state.  These values will outcome in the following states: H, S1, S2, D and staying in the current state, respectively. The prob argument defines the probability of these final_outcomes.  For example, pRecover() defins the probability of recovering and it is a function of state because this probability is only applied if the state is "S1", and is 0 otherwise. Note here we use `Inf` here is as a placeholder to reduce the amount of typing required. Then, we define the payoffs of cost and utility and define each as a function of the state and the decision taken.

The `"curr_state"` is a build-in function that instructs gmod to compute the probability of a state returning to itself.

Before we can evaluate the model, first we need to pass the parameter values. Here we use the global environment to define those. Later, we show how to pass these as a parameter list `param` which will be helpful for analyses such as probabilitistic sensitivity analyses. 
```{r}
params <- list(
          r_HD = 0.002,
          r_HS1 = 0.15,
          r_S1H = 0.5,
          r_S1S2 = 0.105,
          hr_S1 = 3,
          hr_S2 = 10,
          hr_S1S2_trtB = 0.6,
          
          c_H = 2000,
          c_S1 = 4000,
          c_S2 = 15000,
          c_D = 0,
          c_trtA = 12000,
          c_trtB = 13000,
          
          u_H = 1,
          u_S1 = 0.75,
          u_S2 = 0.5,
          u_D = 0,
          u_trtA = 0.95
)
list2env(params, envir = .GlobalEnv)
```


The model has six user-defined functions: four for computing probabilities `pRecover()`, `pGetSick()`, `pProgress()` and `pDie()`, and 2 for payoffs `cost()` and `utility()`.  

probability of recover is only dependent on state being S1 otherwise is 0.  Following the tutorial, we convert the rate of recover r_S1H to a probability by using the function rate2prob().
```{r}
pRecover <- function(state){
  rRecover <- if (state=="S1") r_S1H else 0
  rate2prob(rRecover)
}
```
This is a standard function and can be evaluated like any other function.  For example, we can type
```{r}
#pRecover("S1")

#pRecover("S2")
```

probability of getting sick is also a function of a single state "H" otherwise is 0.
```{r}
pGetSick <- function(state){
  rGetSick <- if (state=="H") r_HS1 else 0
  rate2prob(rGetSick)
}
```

probability of progress is slightly more complex as it is a function of both the state and the decision.  
```{r}
pProgress <- function(state, decision){
  hr_S1S2 <- if (decision %in% c("StrategyB", "StrategyAB")) hr_S1S2_trtB else 1
  rProgress <- (state=="S1") * r_S1S2 * hr_S1S2
  rate2prob(rProgress)
}
```
hr_S1S2 is a function of the decision if it contains "B" then it is hr_S1S2_trtB otherwise is 1.  then the rate of progression is only conditional on state being S1 which gets multiplied by the r_S1S2 and the hr_S1S2.

Using expressions such as (state=="S1") are very helpful as they can reduce the typing required and are easy to read once one gets using to them.  This expression is shorthand of a full if...then statement: 

```{r}
state <- "S1"
# these two statements are equivalent and both return TRUE:
(state=="S1")
if (state=="S1") TRUE else FALSE

```

Probability of death is a function of the state. For states S1 and S2 the rate of mortality is relative to the healthy rate multiplied by a hazard ratio for each state.
```{r}
pDie <- function(state){
  rDie <- switch(state, 
         "H" = r_HD,
         "S1" = r_HD*hr_S1,
         "S2" = r_HD*hr_S2,
         "D" = Inf) #translates to probability of 1
  rate2prob(rDie)
}
```

The payoffs are a cost and utility function. Both are dependent on both the state and the decision. For example only those in S1 and S2 receive treatment.
```{r}
cost <- function(state, decision){
  # cost of decision is only applied if the state is either S1 or S2
  if (state %in% c("S1","S2")){
    c_decision <- switch(decision,
                        "StandardOfCare" = 0,
                        "StrategyA" = c_trtA ,
                        "StrategyB" = c_trtB,
                        "StrategyAB" = c_trtA + c_trtB)}
  else {
    c_decision <- 0
  }
  # cost of the state is a function of the state
  c_state <- switch(state, 
         "H" = c_H,
         "S1" = c_S1,
         "S2" = c_S2,
         "D" = c_D)
  # combine both
  return(c_decision + c_state)
}
```

Similarly the utility function is a function of the state and decision. For H, S2 and D, the utility is independent of the decision. But, Strategies involving "A" improve the utility of S1 from u_S1 to u_trtA. One way to code this is:
```{r}
utility <- function(state, decision){
  u_state <- switch(state, 
         "H" = u_H,
         "S1" = if (decision %in% c("StrategyA", "StrategyAB")) u_trtA else u_S1,
         "S2" = u_S2,
         "D" = u_D)
}
```

Then, we build the gmod Markov model by using the `gmod_gen_model_function()` function.
The output is the structure of the model including the formulae.  

To evaluate the functions, we run `gmod_evaluate_functions()` which returns a set of datasets evaluating the functions. 

```{r}
gmod_evaluate_functions(mygmod)
```

```{r}
model_res <- my_markov_model(mygmod, n_cycles = 75, params = params, return_trace = T)
```

The warning just indicates that no parameters were passed, so gmod uses the values in the global environment.  For analyses, such as probabilistic sensitivity analysis (PSA), we can pass the parameters as a list `param`. 

We can load individual outcomes here that we are interested in. For example, if we are interested in the Markov Trace of the Standard of Care treatment, we can type: 
```{r}
model_res$Trace
rowSums(model_res$Trace)
```

Similarly, we can load the summary outcomes of the costs and utility of the strategies, such that:
```{r}
model_res$summary
```


